### 2012-07-13 Doc/VRIPHYS2012/vriphys2012.tex
changing from SCA 2012 to Vriphys 2012
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@420 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
more on the article
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@422 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\documentclass{egpubl}
\usepackage{sca2012}

@@

% for including postscript figures
% mind: package option 'draft' will replace PS figure by a filname within a frame
\ifpdf \usepackage[pdftex]{graphicx} \pdfcompresslevel=9
\else \usepackage[dvips]{graphicx} \fi

@@

\newcommand{\Figure}[3]{%
\begin{figure}[htb]
  \centering
  \includegraphics[width=#1]{#2}
  \caption{\label{fig-#2}#3}
\end{figure}}

@@

%-------------------------------------------------------------------------
\section{Introduction}

Objects with thin structure form a very specific field of research. Because
they have high ratio between width and thickness methods from volumetric
deformation modeling cannot be immediately used. Such object occur in many
places e.g. textiles, paper or leaves. They are also very common in
anatomical structures for example in form of tubular objects like blood
vessels or colon, in the form of thin membranes, but also many others
(lense, skin, \ldots).

Methods based on continuum mechanics and namely theory of elasticity gained
on popularity ever since Terzopoulos \cite{Terzopoulos1987} presented his
work on elastic deformation modeling. While the methods using finite
elements and theory of elasticity are fairly popular in volumetric modeling
\footnote{TODO: do we need ref. here?} of deformations, they are still
frowned uppon in the area of thin structures. Instead either the
mass-spring models \cite{Volino2009} or bending models
\cite{Grinspun2003,Choi2007} are preffered. Unfortunately they have the
disadvantage that physical parameters of a material (like Young's modulus
and Poisson's ratio) cannot be used directly.

The usual argument agains FEM based methods being that they are too slow.
This is indeed true if lots of elements are used to discretize the object.
However, exploiting the bending property of the shell elements allows us to
achieve good results even with relatively small number of elements. We can
use high-polygonal mesh only for rendering or collision detection and
response. Similar idea has been already presented in the work of Bouthors
et al. \cite{Bouthors2007}. They used twinned meshes for animation of mesh
defined by implicit surface. Simple low-polygonal mesh was used for
mechanical FEM model and the mesh was subdivided in areas with high
curvature after every step of the simulation to produce finer mesh for
rendering. The underlaying FEM model is however not aware of the implicit
surface used. We build uppon the work of Comas \cite{Comas2010c} who used a
simple cubic polynomial to describe the element geometry and derived the
shell model from it.

The Comas's element has several drawbacks that we needed to address. Namely
it produces discontinuities between the elements on the edges, it is
missing one rotationonal degree of freedom and is not completly symetric in
context of deformations.

% TODO: cite Ubach & Oñate?

@@

\section{Preliminaries}

@@

where $P_{i,j}$ are the control points and $B^n_{i,j}$ are the bivairate
Bernstein basis functions defined as:

@@

where $ 0 \le i+j \le n $.

@@

where $ \xi_3 = 1 - \xi_1 - \xi_2 $. By naming the respective values of
the Bernstein basis functions $n_i$ we can shortly express
\eqref{eq-cubicbez} as:

@@

\subsection{Mechanics}

@@

where $U_1, U_2, U_3$ are the translational displacment vectors of the
corner nodes of the element, $\theta_1, \theta_2, \theta_3$ are the angular
displacements of the corner nodes and $P_{i-j}$ are the segments from
corner nodes to other nodes of the control mesh:

@@

and from Kirchhoff-Love theory for thin plates:

@@

where $M$ is the material matrix. To keep the system simple we use linear
Hooke's law for isotropic materials.

@@

The Cauchy's strain tensor \eqref{eq-cauchy} is know to not be rotation
invariant \footnote{add ref}. To deal with rigid body rotations we compute
the displacements and forces in corotational frame. At each step of the
simulation we compute the local frame for every element from the position
of the corner nodes and we allign one axis of the frame with one edge of
the element. \footnote{add image}

The resulting internal force of deformed elemnt is then computed relative
to this frame:

@@

@@

\begin{eqnarray}
    F_1 & = & F \sum_{i \in \{1,4,5\}} n_i + \frac{1}{3} F n_{10} \\
    F_2 & = & F \sum_{i \in \{2,6,7\}} n_i + \frac{1}{3} F n_{10} \\
    F_3 & = & F \sum_{i \in \{3,8,9\}} n_i + \frac{1}{3} F n_{10} \\
\end{eqnarray}

@@

Similarily we compute the torques applied through edge control points and
the central control points by:

\begin{eqnarray}
    \theta_1 & = & P_{1-4} \times (n_4 F) + P_{1-5} \times (n_5 F) + \\
        & + & \frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F) \\
    \theta_2 & = & P_{1-6} \times (n_6 F) + P_{1-7} \times (n_7 F) + \\
        & + & \frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
    \theta_3 & = & P_{1-8} \times (n_8 F) + P_{1-9} \times (n_9 F) + \\
        & + & \frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
\end{eqnarray}

@@

\section{Results}

proof of symmetry

some other examples?

computation speed

\section{Discussion}

% = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
%
% Benefits:
% + C^0 continuity on the edges
% + Perfectly symmetric results for the element (it is not the case with
%   "classical shells")
% + Better conditioning (comparison between Comas model and ours in term of
%   convergence of CG...) * Good trade-off between continuity of interpolation
%   and spasticity: only 18 Dofs per elements for an interpolation with 3d
%   order
% + polynomials [Bezier triangle has theoretically 10 points x 3 dofs=> 30 dofs]
%
% Limitations:
% + Continuity C1 at the nodes but C0 on the edges
% + The segments are at fixed distance from corner nodes, may lead to
%   uneven curvature for too stretched or shrinked elements 
%
% Applications:
% + cataract surgery
% + arterial vessel (MICCAI)
% + vagina/uterus?
%
% = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =

%Long captions should be set as in Figure~\ref{fig:ex1}.
%
%\begin{figure}[htb]
%   % an empty figure just consisting of the caption lines
%   \caption{\label{fig:ex1}
%     'Empty' figure only to serve as an example of long caption requiring 
%     more than one line. It is not typed centered but aligned on both sides.}
%\end{figure}
%
%All graphics should be centered.
%
%%%
%%% Figure 1
%%%
%\begin{figure}[htb]
%  \centering
%  % the following command controls the width of the embedded PS file
%  % (relative to the width of the current column)
%  \includegraphics[width=.8\linewidth]{sampleFig}
%  % replacing the above command with the one below will explicitly set
%  % the bounding box of the PS figure to the rectangle (xl,yl),(xh,yh).
%  % It will also prevent LaTeX from reading the PS file to determine
%  % the bounding box (i.e., it will speed up the compilation process)
%  % \includegraphics[width=.95\linewidth, bb=39 696 126 756]{sampleFig}
%  %
%  \parbox[t]{.9\columnwidth}{\relax
%           For all figures please keep in mind that you \textbf{must not}
%           use images with transparent background! 
%           }
%  %
%  \caption{\label{fig:firstExample}
%           Here is a sample figure.}
%\end{figure}

%If your paper includes images, it is very important that they are of
%sufficient resolution to be faithfully reproduced.
%
%To determine the optimum size (width and height) of an image, measure
%the image's size as it appears in your document (in millimeters), and
%then multiply those two values by 12. The resulting values are the
%optimum $x$ and $y$ resolution, in pixels, of the image. Image quality
%will suffer if these guidelines are not followed.
%
%Example 1: 
%%
%An image measures 50\,mm by 75\,mm when placed in a document. This
%image should have a resolution of no less than 600 pixels by 900
%pixels in order to be reproduced faithfully.

%%------------------------------------------------------------------------
%\subsection{Copyright forms}
%
%You must include your signed Eurographics copyright release form
%when you submit your finished paper. We MUST have this form before
%your paper can be published in the proceedings.

@@

\bibliography{sca2012.bib}

@@

>>>>

\documentclass{egpubl}
\usepackage{vriphys2012}

@@

% for including postscript figures
% mind: package option 'draft' will replace PS figure by a filename within a frame
\ifpdf \usepackage[pdftex]{graphicx} \pdfcompresslevel=9
\else \usepackage[dvips]{graphicx} \fi

@@

%\Figure{width}{name}{caption}
\newcommand{\Figure}[3]{%
\begin{figure}[htb]
  \centering
  \includegraphics[width=#1]{#2}
  \caption{\label{fig-#2}#3}
\end{figure}}

@@

\section{Introduction} % {{{

Objects with thin structure form a very specific field of research. Because
they have high ratio between width and thickness methods from volumetric
deformation modeling cannot be immediately used. Such object occur in many
places e.g. textiles, paper or leaves. They are also very common in
anatomical structures for example in form of tubular objects like blood
vessels or colon, in the form of thin membranes, but also many others
(lens, skin, \ldots).

Methods based on continuum mechanics and namely theory of elasticity gained
on popularity ever since Terzopoulos \cite{Terzopoulos1987} presented his
work on elastic deformation modeling. While the methods using finite
elements and theory of elasticity are fairly popular in volumetric modeling
\footnote{TODO: do we need ref. here?} of deformations, they are still
frowned upon in the area of thin structures. Instead either the
mass-spring models \cite{Volino2009} or bending models
\cite{Grinspun2003,Choi2007} are preferred. Unfortunately they have the
disadvantage that physical parameters of a material (like Young's modulus
and Poisson's ratio) cannot be used directly.

The usual argument against FEM based methods being that they are too slow.
This is indeed true if lots of elements are used to discretize the object.
However, exploiting the bending property of the shell elements allows us to
achieve good results even with relatively small number of elements. We can
use high-polygonal mesh only for rendering or collision detection and
response. Similar idea has been already presented in the work of Bouthors
et al. \cite{Bouthors2007}. They used twinned meshes for animation of mesh
defined by implicit surface. Simple low-polygonal mesh was used for
mechanical FEM model and the mesh was subdivided in areas with high
curvature after every step of the simulation to produce finer mesh for
rendering. The underlaying FEM model is however not aware of the implicit
surface used. We build upon the work of Comas \cite{Comas2010c} who used a
simple cubic polynomial to describe the element geometry and derived the
shell model from it.

The Comas's element has several drawbacks that we needed to address. Namely
it produces discontinuities between the elements on the edges, it is
missing one rotational degree of freedom and is not completely symmetric in
context of deformations.

% TODO: cite Ubach & Oñate
TODO: cite \cite{Ubach2010}: rotation-free shell elements, also based on
Bézier triangle

@@

\section{Preliminaries} % {{{

@@

\noindent
where $P_{i,j}$ are the control points and $B^n_{i,j}$ are the bivairate
Bernstein basis functions defined as:

@@

\noindent
where $ 0 \le i+j \le n $.

@@

\noindent
where $ \xi_3 = 1 - \xi_1 - \xi_2 $. By naming the respective values of
the Bernstein basis functions $n_i$ we can shortly express
\eqref{eq-cubicbez} as:

@@

\subsection{Mechanics} % {{{

@@

\noindent
where $U_1, U_2, U_3$ are the translational displacment vectors of the
corner nodes of the element, $\theta_1, \theta_2, \theta_3$ are the angular
displacements of the corner nodes and $P_{i-j}$ are the segments from
corner nodes to other nodes of the control mesh:

@@

\noindent
and from Kirchhoff-Love theory for thin plates:

@@

\noindent
where $M$ is the material matrix. To keep the system simple we use linear
Hooke's law for isotropic materials.

@@

\subsection{Corotational Formulation} % {{{
\label{sec-corot}

It is known that the Cauchy's strain tensor \eqref{eq-cauchy} is not
rotation invariant \footnote{add ref} and produces ghost forces for rigid
rotations. The alternative approach is to use Green's strain tensor,
however it leads to nonlinear system. To keep the system linear and to deal
with rigid body rotations we compute the displacements and forces in
corotational frame. At each step of the simulation we compute the local
frame for every element, the resulting internal force of deformed element
is then computed relative to this frame:

@@

TODO: how we compute frame?
%from the position of the corner nodes and we align one axis of the frame with one
%edge of the element. \footnote{add image}

%The displacement field defined in \eqref{eq-bezU} is clearly symmetric in
%respect to deformations, but ...

This choice of the frame aligns one axis of the frame with one of the edges
of the triangle. Correct choice of the corotational frame is a tricky
problem \cite{Felippa2005} and incorrect frame for the membrane element can
have adverse effects on the simulation. In our case we have experienced
nonnegligible difference in deformation of equilateral triangle depending
on whether an axis was aligned to the edge favored by the deformation or to
one of the two other edges.

Performing a polar decomposition on the deformation gradient $F = \nabla u
+ I$ can be used to extract a rotation component from the deformation of
the element \cite{Ciarlet1994}. Because we are only fixing the membrane
element we can perform the polar decomposition in 2D with the following
formula:

\begin{equation}
    R = F + \mathrm{sgn}(\mathrm{det}(F)) \begin{pmatrix}
        F_{22} & - F_{21} \\
        F_{12} & F_{11}
    \end{pmatrix}
\end{equation}

\noindent
and subsequently normalizing the columns. We then rotate the corotational
frame in inverse direction. Even single step of this process improves the
final frame but more iterations of the polar decompositions are usually
necessary (up to 20) which severely degrades the performance. But polar
decomposition tends to "overshoot" the ideal rotation angle and each
subsequent step performs a rotation in the opposite direction. By properly
scaling the rotation angle we are able to minimize the amount of needed
iterations down to 5. The best value for the scaling factor depends on the
simulated problem, but we have experimentally localized the best value to
be somewhere between 0.6 and 0.65. We choose to use the value 0.61.

We employ two stop conditions for the iteration process:
\begin{itemize}
    \item The change in rotation angle is less than $10^{-6}$,
    \item and the maximum number of 5 iterations.
\end{itemize}

If the first condition is not met in 5 iterations it can mean one of three
things:

\begin{enumerate}
    \item We still have very good approximation of the ideal frame.
    \item The convergence for the element is too slow.
    \item The process is diverging.
\end{enumerate}

\noindent
and performing any more iterations is unnecessary.

% TODO: maybe make a short table showing the convergence to the symmetric
% solution.

@@

\begin{eqnarray}
    F_1 & = & F \sum_{i \in \{1,4,5\}} n_i + \frac{1}{3} F n_{10} \\
    F_2 & = & F \sum_{i \in \{2,6,7\}} n_i + \frac{1}{3} F n_{10} \\
    F_3 & = & F \sum_{i \in \{3,8,9\}} n_i + \frac{1}{3} F n_{10}
\end{eqnarray}

@@

Similarly we compute the torques applied through edge control points and
the central control points by:

\begin{gather}
    \begin{split}
        \theta_1 & = P_{1-4} \times (n_4 F) + P_{1-5} \times (n_5 F) + \\
        & + \frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F)
    \end{split}
    \\
    \begin{split}
        \theta_2 & = P_{1-6} \times (n_6 F) + P_{1-7} \times (n_7 F) + \\
        & + \frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
    \end{split}
    \\
    \begin{split}
        \theta_3 & = P_{1-8} \times (n_8 F) + P_{1-9} \times (n_9 F) + \\
        & + \frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
    \end{split}
\end{gather}

@@

\section{Validation and Results} % {{{

% We cannot show results for simple plane in bending because the solution
% seems to converge to wrong value (but it converges! :)). Thus showing
% results for plain membrane deformations doesn't make sense (it would be
% inconsistent).

In next two subsections we present tests to validate the convergence of our
Bézier shell element. The results are compared with the shell element
composed of DKT element and optimal ANDES element \cite{Felippa2003}. The
last subsection presents evaluation of the computational complexity of our
solution.

\subsection{Roof Test}

To validate the results of the element we have performed a modified version
of a test known as Scordelis-Lo roof. It simulates a cylindrical roof under
self-weight. The geometry is defined by 80° cylindrical patch of length
$L=50$ and radius $r=25$ and thickness $t=0.25$ and is discretized into
mesh of NxN nodes. The physical parameters are $E = 4.32 \times 10^8$ and
$\nu = 0$. The roof is loaded with uniformly distributed load $q = 90$ per
unit area. The curved edges are clamped at both ends to avoid rigid body
movement. This the only difference from the original Scordelis-Lo roof test
where the edges are free to move in longitudinal direction. The Figure
\ref{fig-roof} shows the vertical displacement of midpoint on the free
edge.

\Figure{\linewidth}{roof}
{Convergence for modified Scordelis-Lo roof test of DKT+ANDES element and
our Bézier shell element.}

\subsection{Hemisphere Test}

Second performed test is a hemisphere with hole subjected to two opposing
forces at the base. The geometry is described by a hemisphere with radius
$r=10$ and thickness $t=0.04$ with 18° hole at the top and the geometry is
discretized into the grid of NxN nodes per quadrant. The physical
parameters are $E = 6.825 \times 10^7$ and $\nu = 0.3$. The hemisphere is
at it's base subjected to two opposing outwards forces and two opposing
inwards forces with magnitude $P = 4$. To avoid rigid body movement we have
constrained two opposite nodes at the top of the hemisphere. The radial
displacement for one point with applied load is shown in Figure
\ref{fig-hemisphere}.

\Figure{\linewidth}{hemisphere}
{Convergence for test on hemisphere with 18° hole for the DKT+ANDES element and
our Bézier shell element.}

\subsection{Computation Speed}

In figure \ref{fig-fps} we present performance of the element in terms of
frames per second. All tests were performed on machine equipped Dual-Core
AMD Opteron Processor 2218 and 3 GB of RAM and conjugate gradient solver
has been to solve the system. Our implementation is without optimizations
and isn't parallel, which means only single CPU core was used for the
simulation. The values reported are for the raw physical simulation with
and without the frame fixing method described in subsection
\ref{sec-corot}. We also present results including the mapping of
high-resolution mesh onto the physical mesh for visualisation purposes or
collisions.

\Figure{\linewidth}{fps}
{Performance in frames per seconds for different number of elements. Shows
values for element with frame fixing (B) method described in subsection
\ref{sec-corot} and without it (A), and with high-resolution mesh of 2048
(C) or 3200 (D) triangles mapped onto the physical mesh.}

We can see that the frame fixing technique is really fast and we are able
to simulate up to 1000 elements and still maintain visual refresh rate of
25 FPS. Our implementation of mechanical mapping clearly needs improving
because it consumes most of the computation time, but it is still able to
provide good results for more than 600 elements.

\section{Discussion and Limitations} % {{{

The element has been successfully tested in SOFA framework for surgical
simulations. The applications being for example simulation pre-operative
planning of cardiac surgery on arterial vessels \cite{Kislinskiy2012}, or
{\em something, something} of vagina/uterus.

% = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
%
% TODO
% Other Applications:
% + cataract surgery
%

Currently, due to the technical decisions the element has two limitations.
The first limitation is that the element does not provide $C^1$ continuity
across edges. It is $C^1$ continuous at corners but only $C^0$ on the
edges. This may prove to be an issue for example if curvature needs to be
computed not only inside the element but also on the edges.

The second limitation is the choice of rigidly attached edge control
points. While it greatly simplifies the formulation the rigidity may cause
issues. In case of large compression of the element the results may be
unpredictable. This is however unlikely to occur because the element will
bend before reaching such configuration. It may, however, fail to maintain
even curvature of the surface if the element is subjected to large
stretching.

On the other hand the choices provide a good trade-off between continuity
and simplicity. Only $ 3 \times 6~\mathrm{DOFs} = 18~\mathrm{DOFs}$ are
necessary while the Bézier triangle has theoretically $ 10 \times
3~\mathrm{DOFs} = 30~\mathrm{DOFs}$.

\subsection{Conclusion} % {{{

This paper presents new shell element with geometry and displacement
function based on the formulation of Bézier triangle. It fixes problems
of the previous model \cite{Comas2010c}, namely:

    \item It is continuous on the edges and provides $C^1$ continuity in
    corners of the element. This is essential for good visual experience if
    high-resolution mesh is mapped on the physical mesh.

    \item Deformations of the element are symmetric. A method of improving
    the corotational frame to keep the deformation symmetric is also
    presented in the paper.

    \item TODO: anything else \ldots? % TODO

    % TODO: check this:
    % + Better conditioning (comparison between Comas model and ours in term of
    %   convergence of CG...)

Because of the formulation of Bézier triangle that is intrinsically present
there is a clean and obvious way of mapping high-resolution mesh for
better visualisation. We also provide a mapping of forces and velocities
between simulation mesh and physical mesh for collision detection/response
and interaction with the mesh which are essential component for the surgical
simulator.

Finally examples showing the convergence to the optimal solution were also provided.

@@

\bibliography{vriphys2012.bib}

@@

% vim: fdm=marker tw=75
